---
title: "Appendix 1: Introduction to R  Control flow"
output: html_notebook
editor_options:
  chunk_output_type: console
---

# Lecture outline

* if-else statements
* for loops
* while loops

# Control flow

Control flow and loops are programming constructs that allow you to execute different code blocks depending on certain conditions or to repeat a block of code a certain number of times.

In R, there are several ways to implement control flow and loops. The three key control flow structure are if-else statements, for loops and while loops.

## If-else statements

If-else statements allow you to execute different blocks of code depending on a condition. Here's an example:

```{r}
# define a function to check if a number is positive or negative
check_sign <- function(x) {
  if (x > 0) {
    print("The number is positive.")
  } else if (x < 0) {
    print("The number is negative.")
  } else {
    print("The number is zero.")
  }
}

# call the function with different arguments
check_sign(5)
# output: The number is positive.

check_sign(-2)
# output: The number is negative.

check_sign(0)
# output: The number is zero.
```

In this example, the `check_sign` function takes one argument, `x`, which represents the number to be checked. The function uses an if-else statement to print a message depending on the value of `x`.

## For loops

For loops allow you to repeat a block of code a certain number of times. Here's an example:

```{r}
# define a function to print the first n even numbers
print_even_numbers <- function(n) {
  for (i in 1:n) {
    print(2*i)
  }
}

# call the function with n = 5
print_even_numbers(5)
# output: 2 4 6 8 10
```

In this example, the `print_even_numbers` function takes one argument, `n`, which represents the number of even numbers to be printed. The function uses a for loop to iterate from 1 to `n` and print the corresponding even numbers.

## While loops

While loops allow you to repeat a block of code as long as a certain condition is true. Here's an example:

```{r}
# define a function to calculate the factorial of a number
factorial <- function(n) {
  result <- 1
  i <- 1

  while (i <= n) {
    result <- result * i
    i <- i + 1
  }

  return(result)
}

# call the function with n = 5
factorial(5)
# output: 120
```

In this example, the `factorial` function takes one argument, `n`, which represents the number whose factorial is to be calculated. The function uses a while loop to calculate the factorial by multiplying the current result by the current value of `i`, which is incremented in each iteration, until `i` reaches `n`.

# Exercises

## If-else

1. Write a function `sign_of_difference` that takes two numbers as arguments and returns "positive" if the difference between them is positive, "negative" if the difference is negative, and "zero" if the difference is zero.

2. Write a function `grade` that takes a numeric score as an argument and returns "A" if the score is greater than or equal to 90, "B" if the score is greater than or equal to 80, "C" if the score is greater than or equal to 70, "D" if the score is greater than or equal to 60, and "F" otherwise.

## For loop

1. Write a function `print_multiples` that takes two numbers as arguments, `n` and `m`, and prints the first `m` multiples of `n`.

2. Write a function `sum_of_squares` that takes a vector of numbers as an argument and returns the sum of the squares of those numbers.

## While loop

1. Write a function `factorial` that takes a positive integer as an argument and returns its factorial.

2. Write a function `even_sum` that returns the sum of all even numbers between 1 and 100 using a while loop.

# Solutions

## If-else statement solutions

1. Solution for `sign_of_difference`:

```{r}
sign_of_difference <- function(x, y) {
  diff <- x - y

  if (diff > 0) {
    return("positive")
  } else if (diff < 0) {
    return("negative")
  } else {
    return("zero")
  }
}

# Example usage:
sign_of_difference(10, 5)
# Output: "positive"
sign_of_difference(5, 10)
# Output: "negative"
sign_of_difference(5, 5)
# Output: "zero"
```

2. Solution for `grade`:

```{r}
grade <- function(score) {
  if (score >= 90) {
    return("A")
  } else if (score >= 80) {
    return("B")
  } else if (score >= 70) {
    return("C")
  } else if (score >= 60) {
    return("D")
  } else {
    return("F")
  }
}

# Example usage:
grade(95)
# Output: "A"
grade(82)
# Output: "B"
grade(75)
# Output: "C"
grade(63)
# Output: "D"
grade(45)
# Output: "F"
```

## For loop solutions

1. Solution for `print_multiples`:

```{r}
print_multiples <- function(n, m) {
  for (i in 1:m) {
    print(n * i)
  }
}

# Example usage:
print_multiples(3, 5)
# Output:
# 3
# 6
# 9
# 12
# 15
```

2. Solution for `sum_of_squares`:

```{r}
sum_of_squares <- function(x) {
  result <- 0

  for (num in x) {
    result <- result + num^2
  }

  return(result)
}

# Example usage:
sum_of_squares(c(2, 3, 4))
# Output: 29
```

## While loop solutions

1. Solution for `factorial`:

```{r}
# Define a function to calculate factorial of a number
factorial <- function(n) {
  if (n < 0) {
    return(NULL)
  }

  if (n == 0) {
    return(1)
  }

  result <- 1
  i <- 1

  while (i <= n) {
    result <- result * i
    i <- i + 1
  }

  return(result)
}
```

2. Solution for `even_sum`

```{r}
even_sum <- function(A, B) {
  if (A > B) {
    temp <- A
    A <- B
    B <- temp
  }

  i <- A
  result <- 0

  while (i <= B) {
    if (i %% 2 == 0) {
      result <- result + i
    }
    i <- i + 1
  }

  return(result)
}
```

