---
title: "Appendix 2: Comparing ADF and KPSS Tests"
output: html_notebook
editor_options: 
  chunk_output_type: console 
---

```{r}
library(zoo)
library(tseries)

stocks <- EuStockMarkets |> zoo::as.zoo()

dax <- stocks$DAX
smi <- stocks$SMI
cac <- stocks$CAC
ftse <- stocks$FTSE
```

## 4.3 Comparing ADF and KPSS Tests

The ADF and KPSS tests can be used together to gain more confidence in our conclusions about the stationarity of a time series. Here's how we can interpret the combined results:

1. If ADF rejects the null hypothesis and KPSS fails to reject the null hypothesis, we have strong evidence of stationarity.
2. If ADF fails to reject the null hypothesis and KPSS rejects the null hypothesis, we have strong evidence of a unit root (non-stationarity).
3. If both tests reject their null hypotheses, we may have a series with some other complex form of non-stationarity.
4. If neither test rejects its null hypothesis, the data may not be sufficiently informative to be sure about stationarity.

Let's summarize our findings for the EU Stock Markets data:

```{r summary_table}
# ADF test on original series
adf_results <- lapply(colnames(EuStockMarkets), function(index) {
  EuStockMarkets[, index] |> adf.test(alternative = "stationary")
})

names(adf_results) <- colnames(EuStockMarkets)

# ADF test on first differences
adf_results_diff <- lapply(colnames(EuStockMarkets), function(index) {
  EuStockMarkets[, index] |>
    diff() |>
    adf.test(alternative = "stationary")
})

names(adf_results_diff) <- colnames(EuStockMarkets)

# KPSS test on original series
kpss_results <- lapply(colnames(EuStockMarkets), function(index) {
  EuStockMarkets[, index] |> kpss.test(null = "Level")
})

names(kpss_results) <- colnames(EuStockMarkets)

# KPSS test on first differences
kpss_results_diff <- lapply(colnames(EuStockMarkets), function(index) {
  EuStockMarkets[, index] |>
    diff() |>
    kpss.test(null = "Level")
})

names(kpss_results_diff) <- colnames(EuStockMarkets)

summary_table <- data.frame(
  ADF_Original = sapply(adf_results, function(x) x$p.value),
  KPSS_Original = sapply(kpss_results, function(x) x$p.value),
  ADF_Differenced = sapply(adf_results_diff, function(x) x$p.value),
  KPSS_Differenced = sapply(kpss_results_diff, function(x) x$p.value)
)

print(summary_table)
```

From this summary, we can conclude:

1. For the original series: Both ADF and KPSS tests suggest non-stationarity (ADF fails to reject, KPSS rejects).
2. For the differenced series: Both ADF and KPSS tests suggest stationarity (ADF rejects, KPSS fails to reject).

These results provide strong evidence that the original EU Stock Markets series are non-stationary, but become stationary after differencing. This is a common characteristic of many financial time series.

In practice, this suggests that we should work with the differenced series (i.e., returns rather than price levels) when building forecasting models or conducting further statistical analyses.
